INGRID: Creating Languages in MPS from ANTLR Grammars

### 2016-10-22 chapters/language_definition.tex
running example

\section{Constructing Languages in MPS}
\label{sect:LANGDEF}

@@

@@

We illustrate the main steps of the INGRID method on the example of a simplified XML language, which is defined below (in Figure XY).

\todo{Pouzit jazyk SimpleXML na running exaple.}
\PV{Dalsi veta asi prijde niz?}

We decided to use SimpleXML because it is small enough to fit here, but still can be used to illustrate the main challenges and our solution

Throughout the whole paper, we will be using one example language to illustrate main challenges and our solution.
The language is a simplified version of XML. We call it \emph{SimpleXML}.
We created the grammar of the language by taking the official XML ANTLRv4 grammar\footnote{https://github.com/antlr/grammars-v4/tree/master/xml} and stripped it down to eliminate/avoid some not so interesting (advanced) features such as XML entities.
Below in Figure X, you can find the grammar of the SimpleXML language in the ANTLRv4 notation.

\begin{figure*}[ht]
\centering
\begin{framed}
\begin{alltt}
	\textbf{grammar} \textit{SimpleXML};
	\antlrparserrule{document}  : \antlrparserrule{prolog}? \antlrparserrule{comment}? \antlrparserrule{element} ;
	\antlrparserrule{prolog}    : \antlrliteral{<?xml } \antlrparserrule{attrib}* \antlrliteral{?>} ;
	\antlrparserrule{comment}   : \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
	\antlrparserrule{element}   : \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>} | \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{/>} ;
	\antlrparserrule{attrib}    : \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT} \antlrliteral{"} ;
	\antlrparserrule{content}   : \antlrlexerrule{TEXT} | \antlrparserrule{element} | \antlrparserrule{comment} | \antlrlexerrule{CDATA} ;
	\antlrlexerrule{Name}      : \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
	\textbf{fragment}
	\antlrlexerrule{DIGIT}     : \antlrregex{[0-9]} ;
	\textbf{fragment}
	\antlrlexerrule{NameChar}  : \antlrlexerrule{NameStartChar} | \antlrliteral{-} | \antlrliteral{\_} | \antlrliteral{.} | \antlrlexerrule{DIGIT} ;
	\textbf{fragment}
	\antlrlexerrule{NameStartChar} : \antlrregex{[:a-zA-Z]} ;
	\antlrlexerrule{TEXT}      : \antlrregex{~[<"]*} ;
	\antlrlexerrule{CDATA}     : \antlrliteral{<![CDATA[} \antlrregex{.*?} \antlrliteral{]]>} ;
\end{alltt}
\end{framed}
\caption{SimpleXML grammar}
\end{figure*}

We have colored the grammar so it is easier to find the way around it (to make navigation easier).
The grammar contains following basic elements:

\begin{itemize}
	\item \textbf{ANTLRv4 keywords} -- In black, they are required by the specification.
	\item \antlrparserrule{\textbf{Parser rules}} -- Stated in blue, they are the rules describing language's structure.
		The alternatives (that the element on the left side may break into) are separated by the pipe ($|$) character.
	\item \antlrlexerrule{\textbf{Lexer rules}} -- In yellow, they are rules describing terminal symbols that the parser is matching against the input.
		Terminals are encoded as string values and regular expressions.
		The \textbf{fragment} keyword states that the lexer sub-rule is just a helper rule that brings more clarity into the notation but is not visible in the parser output.
	\item \antlrliteralnoap{\textbf{String values (literals)}} -- In red, always inside of a pair of apostrophes, they are similar to lexer rules.
		They describe a terminal string, not using regular expressions but by an exact string match.
	\item \antlrregex{\textbf{Regular expressions}} -- Colored in green, they describe a string token to be matched using a regular expression with the special ANTLRv4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
	\item \textbf{Rule operators (?+*)} - Left in black, following any element, they quantify the number of occurrences the prepending element can appear in.
		These operators can also be prepended by an asterisk (*), meaning they are non-greedy.
\end{itemize}

>>>>

\section{Creating Language Definition in MPS}
\label{sect:LANGDEF}

@@

We will describe the INGRID method using the example of a simplified XML language that is defined in Figure~\ref{fig:SIMPLEXML}.
The \emph{SimpleXML} language is small but complex enough to be used for illustration of the main challenges and behavior of the proposed algorithms.

\begin{figure*}[ht]
\centering
\begin{framed}
\begin{alltt}
\textbf{grammar} \textit{SimpleXML};
\antlrparserrule{document}  : \antlrparserrule{prolog}? \antlrparserrule{comment}? \antlrparserrule{element} ;
\antlrparserrule{prolog}    : \antlrliteral{<?xml } \antlrparserrule{attrib}* \antlrliteral{?>} ;
\antlrparserrule{comment}   : \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
\antlrparserrule{element}   : \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>} | \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{/>} ;
\antlrparserrule{attrib}    : \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT} \antlrliteral{"} ;
\antlrparserrule{content}   : \antlrlexerrule{TEXT} | \antlrparserrule{element} | \antlrparserrule{comment} | \antlrlexerrule{CDATA} ;
\antlrlexerrule{Name}      : \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
\textbf{fragment}
\antlrlexerrule{DIGIT}     : \antlrregex{[0-9]} ;
\textbf{fragment}
\antlrlexerrule{NameChar}  : \antlrlexerrule{NameStartChar} | \antlrliteral{-} | \antlrliteral{\_} | \antlrliteral{.} | \antlrlexerrule{DIGIT} ;
\textbf{fragment}
\antlrlexerrule{NameStartChar} : \antlrregex{[:a-zA-Z]} ;
\antlrlexerrule{TEXT}      : \antlrregex{~[<"]*} ;
\antlrlexerrule{CDATA}     : \antlrliteral{<![CDATA[} \antlrregex{.*?} \antlrliteral{]]>} ;
\end{alltt}
\end{framed}
\caption{Grammar of the SimpleXML language in the ANTLR v4 notation}
\label{fig:SIMPLEXML}
\end{figure*}

The grammar of SimpleXML contains elements of the kinds that are listed below.
Each color in Figure~\ref{fig:SIMPLEXML} corresponds to a different kind in a way that is indicated by the list item headers.
\begin{itemize}
	\item \textbf{ANTLR v4 keywords} are required by the notation.
	\item \antlrparserrule{\textbf{Parser rules}} describe the structure of a language.
		Alternatives on the right side of a rule are separated by the pipe character ($|$).
	\item \antlrlexerrule{\textbf{Lexer rules}} describe terminal symbols that the parser matches against the input string.
		A terminal symbol can be encoded as a string value or using a regular expression.
		The \textbf{fragment} keyword states that the lexer rule just brings more clarity into the notation but it is not visible in the parser output.
	\item \antlrliteralnoap{\textbf{String literals}}, always written inside of a pair of single quote marks, also represent terminal symbols, but by exact match to a string constant.
	\item \antlrregex{\textbf{Regular expressions}} describe string tokens to be matched against regular expressions with a special ANTLR v4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
\end{itemize}
Elements on the right side of a rule can be annotated with standard EBNF operators (?+*) that specify the allowed number of occurrences.

@@
